---
title: "Intro to<br/>~~Programming~~<br/>Baking"
format: revealjs
---

## What is this?

:::: {.columns}

::: {.column width="50%"}
- butter
- white sugar
- brown sugar
- vanilla extract
- eggs
:::

::: {.column width="50%"}
- flour
- baking soda
- baking powder
- sea salt
- chocolate chips
:::

::::

A delicious allegory or an overdone metaphor?

## What is required to bake cookies?

. . .

### A recipe

- Ingredients
- Equipment
- Technique

These should be well-defined

## Create and order the steps

1. Gather ingredients
1. Measure ingredients
1. Mix ingredients in bowl
1. Place dough on sheet
1. Bake dough

## Step 2: Measure ingredients

:::: {.columns}

::: {.column width="50%"}
With [cup], measure:

- [1] butter
- [1] white sugar
- [1] brown sugar
- [3] flour
- [2] chocolate chips
:::

::: {.column width="50%"}
With [teaspoon], measure:

- [2] vanilla extract
- [1] baking soda
- [0.5] baking powder
- [1] sea salt
:::

::::

Take:

- [2] eggs

## Step 3: Mix ingredients

In bowl [1]:

1. add [flour]
1. add [baking soda]
1. add [baking powder]
1. add [sea salt]
1. mix

## Step 3: Mix ingredients (cont)

In bowl [2]:

:::: {.columns}

::: {.column width="50%"}
1. add [butter]
1. add [white sugar]
1. add [brown sugar]
1. mix (cream)
1. add [eggs]
1. add [vanilla]
1. mix (beat)
:::

::: {.column width="50%"}
8. add bowl [1]
9. mix
10. add [chocolate chips]
11. mix
:::

::::

## Step 4: Place dough on sheet

1. roll dough from bowl [1] into ball
1. place on cookie sheet [1]
1. repeat as space allows

## Step 5: Bake dough

1. preheat oven [1] at [375° F]
1. place cookie sheet [1] in oven [1]
1. wait [8] minutes
1. remove if cookies are [lightly browned]

## Things to consider

- What if we want to double the amount of cookies? Would it be easy to update the recipe?
- We might substitute ingredients
- During the steps, I used brackets to note items or amounts that might commonly change

## Tie it back to programming

A recipe is an algorithm made of pseudocode

We use syntax to translate pseudocode into [R code]

The compiler translates [R code] into machine code

### Mapping concepts

Ingredients == variables

Verbs == functions

Containers == data structures

Repeat/if == control structures

## Variables

A variable holds stored data

It will have a data type

Two common data types are numbers and strings. In R these are called "numeric" and "character".

## Data structures

A data structure is a collection of variables

The simplest data structure in R is a vector

[1, 2, 3, 4, 5] is a vector with five numeric elements

## Functions

Functions allow us to repeat common tasks

They usually have input, called arguments

## Control structures

Control structures allow code chunks to run conditionally

- if condition exists
- repeat until condition exists

## Practice example with R vector {.smaller}

This creates the variable "x", which is a numeric vector of length 1

```{r echo=TRUE}
x <- 1
x = 1
1 -> x
```

This creates the variable "y", which is a numeric vector of length 5

```{r echo=TRUE}
y <- c(1, 2, 3, 4, 5)
y <- 1:5
y <- seq(1, 5)
```

This creates the variable "z", which is a character vector of length 5

```{r echo=TRUE}
z <- c('ant', 'bat', 'cat', 'dog', 'eel')
z <- c("ant", "bat", "cat", "dog", "eel")
```

## Practice (cont) {.smaller}

What do you think this does?

:::: {.columns}

::: {.column width="50%"}
### In R

```{r echo=TRUE, eval=FALSE}
y <- c(1, 2, 3, 4, 5)
y[3]
y[3:4]
y[c(1,3,5)]
```

:::

::: {.column width="50%"}
### In Python

```{python echo=TRUE, eval=FALSE}
y = [1, 2, 3, 4, 5]
y[3]
y[3:4]
y[[1,3,5]]
```
:::

::::

. . .

:::: {.columns}

::: {.column width="50%"}
### In R

```{r}
y[3]
y[3:4]
y[c(1,3,5)]
```

:::

::: {.column width="50%"}
### In Python

```{python}
y = [1, 2, 3, 4, 5]
y[3]
y[3:4]
try:
  y[[1,3,5]]
except TypeError as err:
  print(err)
```
:::

::::

. . .

What about this (x = `r x`)?

```{r echo=TRUE, eval=FALSE}
y[x]
```

. . .

```{r}
y[x]
```

## Practice (cont)

Or this (z = `r z`)?

```{r echo=TRUE, eval=FALSE}
y[c(1,1,3,5,4,4,4,2)]
z[c(1,1,3,5,4,4,4,2)]
```

. . .

```{r}
y[c(1,1,3,5,4,4,4,2)]
z[c(1,1,3,5,4,4,4,2)]
```

. . .

Or even

```{r echo=TRUE, eval=FALSE}
y[3:5][2]
```

. . .

```{r}
y[3:5][2]
```

## Key programming concepts

- generalize
- write functions (DRY)
- be consistent in syntax
- add comments for clarity
- writing code can be experimental and is an iterative process

## R resources {.smaller}

type|site|link
-----|-----|-----
links|stackoverflow|<https://stackoverflow.com/tags/r/info>
forum|stackoverflow|<https://stackoverflow.com/questions/tagged/r>
forum|Rstudio community|<https://community.rstudio.com/>
book|R for Data Science|<http://r4ds.had.co.nz/>
book|Advanced R|<http://adv-r.had.co.nz/>
blog|R-bloggers|<http://www.r-bloggers.com/>
blog|RStudio blog|<https://blog.rstudio.com/>
tutorial|Matloff's GitHub|<https://github.com/matloff/fasteR>

## R documents/scripts {.smaller}

: {tbl-colwidths="[20,50,30]"}

term|description|do you care?
----|----|----
RStudio|IDE, environment for developing R code|yes
LaTeX|typesetting language|no
Rnw|mix of R code with LaTeX|no
md|lightweight markup language|yes
Rmd|mix of R code with markdown|yes
qmd|quarto is next-gen Rmarkdown|yes
knitr|R package that dynamically converts R chunks into structured text|yes
pandoc|converts document format|no
shiny|R code execution through web page|maybe

## R packages

An R package is a collection of code to perform some task.

### Terms to know

- CRAN (install.packages)
- GitHub (devtools::install_github)
- binary (ready for use on given OS)
- source (must compile C/Fortran code)

## Find Practice Problems

- <https://www.r-exercises.com/>
- <https://projecteuler.net/archives>
- <https://www.kaggle.com/competitions>
- <https://fivethirtyeight.com/tag/the-riddler/>

## About directories

The home area is commonly referred to with `~/`

: {tbl-colwidths="[15,25,60]"}

+-----------+----------+-------------------------------+
|System     |Home Area |Path                           |
+===========+==========+===============================+
|Windows    |home<br />|C:\\Users\\username<br />      |
|           |R home    |C:\\Users\\username\\Documents |
+-----------+----------+-------------------------------+
|macOS      |home      |/users/username                |
+-----------+----------+-------------------------------+
|Linux/Unix |home      |/home/username                 |
+-----------+----------+-------------------------------+

Normally when we start R, the working, or, current directory will be set to our home area.

## "Working" directory {.smaller}

Consider the home directory for user `mctest` on a `macOS` system. Assume it has two subdirectories, "project" and "Downloads".

The path to the project directory is `/users/mctest/project`, or `~/project`.

From R we could set the working directory to be our project directory with `setwd('~/project')`.

Our working directory has two special paths:

path|description|location
----|----|----
.|itself|/users/mctest/project
..|parent|/users/mctest

## Absolute and relative paths

If our working directory is `/users/mctest/project`, there are two ways we can access the "Downloads" directory.

- absolute path: `/users/mctest/Downloads`
- relative path: `../Downloads`

What directory would be at `../../`?

. . .

`/users`

<!--
Rscript -e 'quarto::quarto_render("go.qmd")'
quarto render go.qmd
-->
